# Module II

---

::: {.learning-objectives}
**Learning Objectives**

After completing this unit, students will be able to:

1. Explain the meaning, features, and types of shares, debentures, and other financial securities
2. Distinguish between equity and preference shares, and among the types of each
3. Analyse the structure, advantages, and risks of mutual fund investments
4. Calculate Net Asset Value (NAV) and Expense Ratio with numerical illustrations
5. Describe the SEBI categorisation of mutual fund schemes (2017–present)
6. Evaluate Exchange Traded Funds (ETFs), REITs, and InvITs as investment instruments
7. Assess the growth and future prospects of ETFs, REITs, and InvITs in the Indian market
:::

---

## 2.1 Shares — Meaning and Features {#shares-meaning}

### 2.1.1 What is a Share?

A **share** is a unit of ownership in a company. When a company raises capital by issuing shares to the public, it distributes ownership of the company into small, equal units. Each unit is a **share**. The holder of a share is called a **shareholder** or **stockholder** and has a proportionate claim on the company's assets and earnings.

In India, shares are issued under the **Companies Act, 2013** and are regulated by **SEBI** (Securities and Exchange Board of India). They are traded on stock exchanges — primarily the **NSE (National Stock Exchange)** and **BSE (Bombay Stock Exchange)**.

> *"A share is not just a piece of paper — it is a proportionate ownership of a living, breathing business enterprise."* — Benjamin Graham

**Share Capital Structure:**

$$\text{Total Equity} = \text{Paid-up Capital} + \text{Reserves and Surplus}$$
$$\text{Book Value per Share} = \frac{\text{Total Equity}}{\text{Number of Outstanding Shares}}$$

### 2.1.2 Rights of Shareholders

Shareholders enjoy several fundamental rights:

1. **Right to dividend** — proportionate share in distributed profits
2. **Right to vote** — one share, one vote (equity shareholders) on company matters
3. **Right to information** — access to annual reports, quarterly results
4. **Right to transfer** — free transfer of listed shares on stock exchanges
5. **Pre-emptive right (Right Issue)** — first right to subscribe to new share issues
6. **Right to residual claims** — claim on assets during liquidation (after creditors)
7. **Right to attend AGM** — participate in Annual General Meetings

---

## 2.2 Equity Shares {#equity-shares}

### 2.2.1 Meaning and Features

**Equity shares** (also called **ordinary shares** or **common stock**) represent the primary and permanent capital of a company. Equity shareholders are the **true owners** of the business and bear the highest risk.

**Key Features:**
- **Variable dividend** — declared by the Board, not fixed
- **Last in dividend priority** — preference shareholders paid first
- **Residual claimants** — in liquidation, equity holders receive what remains after all liabilities
- **Voting rights** — full voting rights in all company matters
- **Unlimited upside** — shares of companies like Infosys, TCS, and Reliance have created immense wealth
- **Perpetual existence** — equity capital is permanent (no redemption)

### 2.2.2 Types of Equity Shares

```{r equity-types, echo=FALSE}
equity_types <- data.frame(
  Type = c("Ordinary / Common Shares", "Founder Shares / Promoter Shares",
           "Sweat Equity Shares", "Employee Stock Option Plan (ESOP)",
           "Bonus Shares", "Right Shares", "Deferred Shares"),
  Features = c(
    "Standard equity with full voting rights and dividend entitlement",
    "Held by founders; may carry differential voting rights (DVR)",
    "Issued to directors/employees for know-how or IPR at discount",
    "Options to buy shares at predetermined price; reward tool",
    "Issued free to existing shareholders from reserves",
    "Offered to existing shareholders at discount before public issue",
    "Dividend deferred until preference and equity holders are paid"
  ),
  Indian_Example = c(
    "Reliance Industries, TCS, HDFC Bank",
    "Infosys founder shares (N.R. Narayana Murthy, Nilekani)",
    "ESOP-heavy tech startups: Zomato, Nykaa",
    "Wipro, Infosys ESOP programmes",
    "TCS 1:1 bonus (2022), Asian Paints frequent bonuses",
    "Tata Motors Rights Issue (2021 — ₹15,000 Cr)",
    "Historically used; rare in modern India"
  )
)

kable(equity_types,
      caption = "Table 2.1: Types of Equity Shares and Indian Examples",
      col.names = c("Type", "Features", "Indian Example"),
      booktabs = TRUE) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), font_size = 11, full_width = TRUE) %>%
  row_spec(0, bold = TRUE, color = "white", background = "#2c3e50")
```

*Source: Compiled from Companies Act 2013, SEBI Regulations, and NSE/BSE corporate action data.*

**Differential Voting Rights (DVR) Shares:**
SEBI amended regulations in 2019 to allow **superior voting right shares (SR Shares)** for promoters of technology companies during their IPO. This was triggered by concerns that founders of new-age companies (like Zomato, Paytm) would lose control post-IPO to institutional investors. The concept borrows from Google's Class A / Class B share structure.

---

## 2.3 Preference Shares {#preference-shares}

### 2.3.1 Meaning and Features

**Preference shares** occupy a **middle position** between equity shares and debentures. Preference shareholders have a **preferential right** over equity shareholders in:
- Payment of **dividends** (fixed rate)
- **Repayment of capital** in case of winding up

However, preference shareholders generally **do not have voting rights** (unless dividends are in arrear for two consecutive years).

$$\text{Preference Dividend} = \text{Face Value} \times \text{Dividend Rate (\%)}$$

### 2.3.2 Types of Preference Shares

```{r pref-share-types, echo=FALSE}
pref_types <- data.frame(
  Type = c("Cumulative", "Non-Cumulative", "Participating", "Non-Participating",
           "Redeemable", "Irredeemable", "Convertible", "Non-Convertible"),
  Description = c(
    "Unpaid dividends accumulate; must be paid before equity dividend",
    "Lapsed dividends are not carried forward",
    "Right to share in surplus profits after fixed dividend",
    "Only entitled to fixed dividend, no additional profit share",
    "Redeemed by company after a specified period (max 20 yrs in India)",
    "Not redeemed during company's lifetime (banned in India post-1988)",
    "Can be converted into equity shares after specified period",
    "Cannot be converted into equity"
  ),
  Key_Feature = c(
    "Investor-protective; safe during loss years",
    "No backlog; company-friendly",
    "Upside participation in good years",
    "Fixed income only",
    "Capital returned; SEBI regulated",
    "Permanent capital; phased out",
    "Equity upside with preference security",
    "Pure fixed-income; no equity participation"
  )
)

kable(pref_types,
      caption = "Table 2.2: Types of Preference Shares",
      booktabs = TRUE) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), font_size = 11, full_width = TRUE) %>%
  row_spec(0, bold = TRUE, color = "white", background = "#2c3e50")
```

*Source: Compiled from Companies Act, 2013 and SEBI Regulations.*

---

## 2.4 Debentures {#debentures}

### 2.4.1 Meaning

A **debenture** is a debt instrument issued by a company acknowledging its indebtedness. It represents a long-term loan to the company. Debenture holders are **creditors**, not owners, and have a legal right to receive:
- Periodic **interest** (called coupon)
- **Repayment of principal** at maturity

In India, debentures issued to public must be listed on stock exchanges (NSE or BSE debt segment) and rated by credit rating agencies like **CRISIL**, **ICRA**, **CARE**, or **India Ratings**.

### 2.4.2 Types of Debentures

```{r deb-types, echo=FALSE}
deb_data <- data.frame(
  Classification = c("On Security", "On Security", "On Convertibility", "On Convertibility",
                     "On Convertibility", "On Redemption", "On Coupon"),
  Type = c("Secured Debentures", "Unsecured Debentures",
           "Fully Convertible Debentures (FCDs)",
           "Partly Convertible Debentures (PCDs)",
           "Non-Convertible Debentures (NCDs)",
           "Redeemable / Irredeemable",
           "Fixed / Floating Rate"),
  Description = c(
    "Backed by charge on company's assets; safer for investors",
    "No asset backing; higher risk, higher yield",
    "Entire amount converted to equity on specified date",
    "Partial conversion to equity; balance redeemed",
    "Not convertible; pure debt instrument; most common in India",
    "Redeemable at par/premium on maturity; irredeemable rare",
    "Fixed coupon vs. linked to MIBOR/T-Bill rate"
  )
)

kable(deb_data,
      caption = "Table 2.3: Types of Debentures",
      col.names = c("Classification Basis", "Type", "Description"),
      booktabs = TRUE) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = TRUE) %>%
  row_spec(0, bold = TRUE, color = "white", background = "#2c3e50") %>%
  collapse_rows(columns = 1, valign = "middle")
```

*Source: Compiled from Companies Act, 2013 and SEBI (LODR) Regulations, 2015.*

**Indian NCD Market:** Non-Convertible Debentures (NCDs) have become a popular retail investment avenue in India. Bajaj Finance, HDFC, Shriram Transport Finance, and Tata Capital have issued retail NCDs with coupon rates ranging from 7.5% to 10% p.a., offering returns superior to bank FDs with credit risk.

---

## 2.5 Mutual Funds {#mutual-funds}

### 2.5.1 Meaning and Structure

A **mutual fund** is a professionally managed collective investment vehicle that pools money from many investors and invests in a diversified portfolio of securities. In India, mutual funds are regulated by **SEBI** under the **SEBI (Mutual Funds) Regulations, 1996** (amended multiple times through 2023).

```{mermaid}
flowchart TD
    A[Investors / Unit Holders] -->|Pool Money| B[Asset Management Company - AMC]
    B -->|Invests| C[Diversified Portfolio]
    C --> D[Equity Shares]
    C --> E[Government Bonds]
    C --> F[Corporate Debentures]
    C --> G[Money Market Instruments]
    C --> H[Gold / REITs]
    B --> I[Fund Manager]
    I --> B
    B -->|Returns| A
    J[SEBI] -->|Regulates| B
    K[Trustee] -->|Oversees| B
    L[Custodian] -->|Holds Assets| B
    M[Registrar & Transfer Agent] -->|Handles Units| A
    style A fill:#2980b9,color:#fff
    style B fill:#2c3e50,color:#fff
    style J fill:#c0392b,color:#fff
    style I fill:#27ae60,color:#fff
```

*Figure 2.1: Three-tier Structure of Indian Mutual Funds*

**Three-Tier Structure:**
1. **Sponsors** — the parent company that establishes the mutual fund (e.g., HDFC Ltd. → HDFC AMC)
2. **Trustees** — independent board overseeing the AMC, protecting investors' interests
3. **AMC (Asset Management Company)** — manages investments, employs fund managers

**Key Industry Statistics:**

```{r mf-industry-growth, echo=FALSE, fig.cap="Figure 2.2: Indian Mutual Fund Industry AUM Growth (2016–2026)"}
mf_years <- 2016:2026
mf_aum <- c(13.46, 17.54, 22.86, 24.51, 26.33, 31.02,
             37.72, 40.04, 50.78, 58.91, 67.40) # ₹ Lakh Crore (approximate)

mf_df <- data.frame(Year = mf_years, AUM = mf_aum)

ggplot(mf_df, aes(x = Year, y = AUM)) +
  geom_col(fill = "#2980b9", alpha = 0.8) +
  geom_line(aes(group = 1), color = "#c0392b", size = 1.5) +
  geom_point(color = "#c0392b", size = 3) +
  geom_text(aes(label = paste0("₹", AUM, "L Cr")), vjust = -0.5, size = 2.8, color = "#2c3e50") +
  labs(title = "Indian Mutual Fund Industry AUM (2016–2026)",
       subtitle = "Total AUM grew from ₹13.46 Lakh Crore to ₹67.40 Lakh Crore",
       x = "Year", y = "AUM (₹ Lakh Crore)") +
  scale_x_continuous(breaks = 2016:2026) +
  scale_y_continuous(expand = c(0.05, 0)) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(face = "bold", color = "#2c3e50"))
```

*Source: Compiled from RBI, SEBI, NSE, BSE, AMFI and Government Reports (Latest Available Data).*

### 2.5.2 Advantages of Mutual Funds

1. **Professional Management** — Experienced fund managers conduct research and manage portfolios
2. **Diversification** — Even ₹500 SIP invests in 50–100 stocks
3. **Affordability** — Start SIP with as low as ₹100/month
4. **Liquidity** — Open-end funds redeemable any business day at NAV
5. **Transparency** — Daily NAV disclosure; monthly portfolio disclosure
6. **Regulatory Protection** — SEBI regulations protect investors; AMFI code of conduct
7. **Tax Efficiency** — ELSS funds offer Section 80C benefits; long-term capital gains taxed at 10% (>₹1 lakh)
8. **Flexibility** — SIP, SWP, STP options; switch between funds
9. **Economies of Scale** — Lower transaction costs due to bulk purchases
10. **Variety** — 40+ AMCs, 1,500+ schemes across all risk profiles

### 2.5.3 Risks in Mutual Fund Investment

| Risk Type | Description | Affected Funds |
|-----------|-------------|----------------|
| Market Risk | Portfolio value falls with market | Equity funds |
| Credit Risk | Issuer defaults on bond payment | Debt funds |
| Interest Rate Risk | Bond prices fall when rates rise | Debt funds |
| Liquidity Risk | Fund unable to sell assets | Close-end funds |
| Concentration Risk | Overdependence on few stocks/sectors | Thematic funds |
| Currency Risk | Loss from foreign exchange movement | International funds |
| Manager Risk | Fund manager's incorrect decisions | All actively managed funds |
| Inflation Risk | Returns don't beat inflation | Conservative funds |

*Source: Compiled from RBI, SEBI, NSE, BSE, AMFI and Government Reports (Latest Available Data).*

---

## 2.6 Net Asset Value (NAV) {#nav}

### 2.6.1 Meaning of NAV

The **Net Asset Value (NAV)** is the per-unit market value of a mutual fund scheme. It represents the price at which investors can buy (subscription) or sell (redemption) units of the fund.

$$\boxed{NAV = \frac{\text{Total Assets} - \text{Total Liabilities}}{\text{Number of Units Outstanding}}}$$

**Components:**
- **Total Assets** = Market value of securities + Cash and equivalents + Receivables + Accrued income
- **Total Liabilities** = Payables + Accrued expenses + Management fee payable

### 2.6.2 NAV Calculation — Worked Example

**Example 2.1:** Calculate the NAV of "Axis Bluechip Fund" given the following data as on March 31, 2024:

| Component | Amount (₹ Crore) |
|-----------|-----------------|
| Equity portfolio market value | 8,450 |
| Liquid funds / Cash | 320 |
| Dividend receivable | 48 |
| Other receivables | 22 |
| Management fee payable | 15 |
| Other liabilities | 8 |
| Number of units | 250 crore |

**Solution:**
$$\text{Total Assets} = 8,450 + 320 + 48 + 22 = ₹8,840 \text{ Cr}$$
$$\text{Total Liabilities} = 15 + 8 = ₹23 \text{ Cr}$$
$$\text{Net Assets} = 8,840 - 23 = ₹8,817 \text{ Cr}$$
$$NAV = \frac{8,817 \text{ Cr}}{250 \text{ Cr units}} = ₹35.27 \text{ per unit}$$

```{r nav-calculation, echo=TRUE}
# NAV Calculation
total_assets <- 8450 + 320 + 48 + 22  # in Crore
total_liabilities <- 15 + 8
net_assets <- total_assets - total_liabilities
units_outstanding <- 250  # Crore units

NAV <- net_assets / units_outstanding
cat("Total Assets: ₹", total_assets, "Crore\n")
cat("Total Liabilities: ₹", total_liabilities, "Crore\n")
cat("Net Assets: ₹", net_assets, "Crore\n")
cat("NAV per unit: ₹", round(NAV, 2), "\n")
```

### 2.6.3 NAV and Returns

**Absolute Return:**
$$\text{Absolute Return (\%)} = \frac{NAV_{current} - NAV_{purchase}}{NAV_{purchase}} \times 100$$

**CAGR (for periods > 1 year):**
$$CAGR = \left(\frac{NAV_{current}}{NAV_{purchase}}\right)^{\frac{1}{n}} - 1$$

**Example 2.2:** An investor purchased units of HDFC Flexi Cap Fund at NAV ₹18.50 on April 1, 2016. The NAV on March 31, 2024 was ₹52.80. Calculate absolute return and CAGR.

```{r nav-returns, echo=TRUE}
nav_purchase <- 18.50
nav_current <- 52.80
years_held <- 8

absolute_return <- ((nav_current - nav_purchase) / nav_purchase) * 100
cagr <- (nav_current / nav_purchase)^(1/years_held) - 1

cat("Absolute Return:", round(absolute_return, 2), "%\n")
cat("CAGR:", round(cagr * 100, 2), "% per annum\n")
```

---

## 2.7 Expense Ratio {#expense-ratio}

### 2.7.1 Meaning and Importance

The **Expense Ratio (ER)** is the annual cost charged by a mutual fund to manage an investor's money, expressed as a percentage of the fund's daily average AUM. It covers:
- **Fund management fee** (largest component)
- **Trustee fee**
- **Registrar and transfer agent fee**
- **Marketing and distribution expenses** (only in Regular Plans)
- **Audit, legal, and custodian fees**

$$\boxed{\text{Expense Ratio (\%)} = \frac{\text{Total Annual Fund Expenses}}{\text{Average Daily AUM}} \times 100}$$

**SEBI Expense Ratio Limits (as of 2024):**

```{r expense-ratio-limits, echo=FALSE}
er_limits <- data.frame(
  AUM_Slab = c("First ₹500 Cr", "Next ₹250 Cr", "Next ₹1,250 Cr",
               "Next ₹3,000 Cr", "Next ₹5,000 Cr", "Next ₹40,000 Cr",
               "Balance above ₹50,000 Cr"),
  Equity_TER = c("2.25%", "2.00%", "1.75%", "1.60%", "1.50%",
                  "Reduce by 0.05% per ₹5,000 Cr", "1.05%"),
  Debt_TER = c("2.00%", "1.75%", "1.50%", "1.35%", "1.25%",
                "Reduce by 0.05%", "0.80%"),
  Index_ETF = c("1.00%", "1.00%", "1.00%", "1.00%", "1.00%", "1.00%", "1.00%")
)

kable(er_limits,
      caption = "Table 2.4: SEBI Maximum TER Limits for Mutual Funds",
      col.names = c("AUM Slab", "Equity TER Max", "Debt TER Max", "Index/ETF Max"),
      booktabs = TRUE) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = TRUE) %>%
  row_spec(0, bold = TRUE, color = "white", background = "#2c3e50")
```

*Source: Compiled from RBI, SEBI, NSE, BSE, AMFI and Government Reports (Latest Available Data).*

### 2.7.2 Expense Ratio Calculation — Worked Example

**Example 2.3:** A mutual fund has daily average AUM of ₹12,000 Crore. Annual expenses: Management fee ₹108 Cr, Distribution ₹36 Cr, Trustee fee ₹6 Cr, Others ₹10 Cr. Calculate Total Expense Ratio.

```{r expense-ratio-calc, echo=TRUE}
aum <- 12000  # Crore
management_fee <- 108
distribution <- 36
trustee <- 6
others <- 10

total_expenses <- management_fee + distribution + trustee + others
TER <- (total_expenses / aum) * 100

cat("Total Annual Expenses: ₹", total_expenses, "Crore\n")
cat("Average AUM: ₹", aum, "Crore\n")
cat("Total Expense Ratio (TER):", round(TER, 3), "%\n")
cat("Impact on ₹1 Lakh invested:", round(100000 * TER/100), "per annum\n")
```

**Direct vs. Regular Plans:**
SEBI mandated Direct Plans in 2013 — these exclude distributor commissions and hence have **lower expense ratios by 0.5%–1.2%**. Over 20 years, this difference compounds significantly.

```{r direct-vs-regular, echo=TRUE, fig.cap="Figure 2.3: Direct Plan vs. Regular Plan Wealth Creation (20-Year SIP ₹10,000/Month)"}
monthly_sip <- 10000
months <- 240
gross_return <- 0.12 / 12  # Monthly

# Regular plan: 12% gross, 1.5% TER → ~10.5% net return
regular_return <- (0.12 - 0.015) / 12
direct_return <- (0.12 - 0.005) / 12  # Direct: 0.5% TER

fv_regular <- monthly_sip * (((1 + regular_return)^months - 1) / regular_return)
fv_direct <- monthly_sip * (((1 + direct_return)^months - 1) / direct_return)

cat("20-Year SIP in Regular Plan: ₹", format(round(fv_regular), big.mark = ","), "\n")
cat("20-Year SIP in Direct Plan:  ₹", format(round(fv_direct), big.mark = ","), "\n")
cat("Extra wealth from Direct Plan: ₹", format(round(fv_direct - fv_regular), big.mark = ","), "\n")

# Plot comparison
year_seq <- 1:20
corpus_regular <- sapply(year_seq, function(y) {
  m <- y * 12
  monthly_sip * (((1 + regular_return)^m - 1) / regular_return)
})
corpus_direct <- sapply(year_seq, function(y) {
  m <- y * 12
  monthly_sip * (((1 + direct_return)^m - 1) / direct_return)
})

plot_df <- data.frame(
  Year = rep(year_seq, 2),
  Corpus = c(corpus_regular, corpus_direct) / 100000,
  Plan = rep(c("Regular Plan (TER 1.5%)", "Direct Plan (TER 0.5%)"), each = 20)
)

ggplot(plot_df, aes(x = Year, y = Corpus, color = Plan)) +
  geom_line(size = 1.4) +
  geom_point(size = 2) +
  scale_color_manual(values = c("Regular Plan (TER 1.5%)" = "#c0392b",
                                 "Direct Plan (TER 0.5%)" = "#27ae60")) +
  labs(title = "Direct vs. Regular Plan: Impact of Expense Ratio",
       subtitle = "SIP ₹10,000/Month, 20 Years, 12% Gross Return",
       x = "Year", y = "Corpus (₹ Lakhs)", color = "") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```

*Source: Compiled from RBI, SEBI, NSE, BSE, AMFI and Government Reports (Latest Available Data).*

---

## 2.8 Types of Mutual Funds {#mf-types}

### 2.8.1 SEBI Categorisation of Mutual Fund Schemes (2017)

SEBI issued a landmark **circular on mutual fund categorisation and rationalisation in October 2017** (effective June 2018), mandating that each AMC can offer only **one scheme per category**. This brought standardisation and comparability to India's mutual fund landscape.

```{mermaid}
flowchart TD
    A[SEBI Categorisation of Mutual Funds] --> B[Equity Schemes]
    A --> C[Debt Schemes]
    A --> D[Hybrid Schemes]
    A --> E[Solution-Oriented Schemes]
    A --> F[Other Schemes]
    
    B --> B1[Multi Cap / Flexi Cap]
    B --> B2[Large Cap / Mid Cap / Small Cap]
    B --> B3[ELSS - Tax Saving]
    B --> B4[Sectoral / Thematic]
    B --> B5[Focused Fund]
    
    C --> C1[Overnight / Liquid]
    C --> C2[Short / Medium Duration]
    C --> C3[Credit Risk Fund]
    C --> C4[Gilt Fund]
    C --> C5[Corporate Bond Fund]
    
    D --> D1[Aggressive Hybrid]
    D --> D2[Balanced Advantage / Dynamic]
    D --> D3[Multi-Asset Allocation]
    D --> D4[Arbitrage Fund]
    
    E --> E1[Retirement Fund]
    E --> E2[Children's Fund]
    
    F --> F1[Index Funds / ETFs]
    F --> F2[FoF - Fund of Funds]
    
    style A fill:#2c3e50,color:#fff
    style B fill:#2980b9,color:#fff
    style C fill:#27ae60,color:#fff
    style D fill:#e67e22,color:#fff
    style E fill:#8e44ad,color:#fff
    style F fill:#c0392b,color:#fff
```

*Figure 2.4: SEBI Categorisation of Mutual Fund Schemes (2017)*

### 2.8.2 Major Categories — Detailed Analysis

**A. Large Cap Funds**
- Must invest minimum **80% in top 100 companies by market capitalisation** (as defined by AMFI)
- Lower risk than mid/small cap; benchmark: Nifty 50 / Sensex
- Suitable for conservative equity investors
- Examples: Axis Bluechip Fund, Mirae Asset Large Cap Fund, ICICI Prudential Bluechip Fund

**B. Mid Cap Funds**
- Must invest minimum **65% in 101st–250th companies** by market cap
- Higher growth potential; higher volatility
- Examples: HDFC Mid-Cap Opportunities Fund, Kotak Emerging Equity Fund

**C. Small Cap Funds**
- Minimum **65% in 251st company onwards** by market cap
- Highest growth potential and highest risk among equity categories
- Examples: Nippon India Small Cap Fund, SBI Small Cap Fund

**D. ELSS (Equity Linked Savings Scheme)**
- Minimum **80% in equity**
- **3-year lock-in** (shortest among 80C investments)
- Section 80C benefit up to ₹1.5 lakh
- Examples: Axis Long Term Equity, Mirae Asset Tax Saver Fund

**E. Liquid Funds**
- Invest in money market instruments with maturity up to **91 days**
- Very low risk; returns slightly above savings account (5.5–6.5% p.a.)
- Suitable for emergency fund management

**F. Gilt Funds**
- Invest **80% minimum in government securities**
- Zero credit risk; high interest rate risk
- Suitable when interest rates are expected to fall

```{r mf-category-comparison, echo=FALSE, fig.cap="Figure 2.5: Mutual Fund Category Risk-Return Matrix (2016–2026)"}
mf_categories <- data.frame(
  Category = c("Liquid", "Overnight", "Short Duration", "Corporate Bond",
               "Gilt", "Balanced Hybrid", "Large Cap", "Flexi Cap",
               "Mid Cap", "Small Cap", "ELSS", "Sectoral"),
  Risk = c(1, 1, 2, 3, 4, 5, 6, 7, 8, 9, 7, 9),
  Return_3yr = c(6.2, 6.0, 6.8, 7.5, 8.5, 11.0, 13.2, 15.8, 18.5, 21.2, 14.5, 20.0)
)

ggplot(mf_categories, aes(x = Risk, y = Return_3yr, label = Category)) +
  geom_point(aes(color = Return_3yr), size = 5) +
  ggrepel::geom_text_repel(size = 3.2, fontface = "bold") +
  scale_color_gradient(low = "#27ae60", high = "#c0392b") +
  labs(title = "Mutual Fund Category: Risk vs. 3-Year Return",
       subtitle = "Indian Mutual Funds (Approximate 3-Year CAGR, 2021–2024)",
       x = "Risk Level (1 = Lowest, 9 = Highest)",
       y = "3-Year CAGR (%)", color = "Return (%)") +
  theme_minimal(base_size = 12)
```

*Source: Compiled from RBI, SEBI, NSE, BSE, AMFI and Government Reports (Latest Available Data).*

**SIP Growth in India (2016–2026):**

```{r sip-growth-india, echo=FALSE, fig.cap="Figure 2.6: SIP Monthly Contribution and Folios (2016–2026)"}
sip_data <- data.frame(
  Year = 2016:2026,
  SIP_Monthly_Cr = c(3122, 4571, 7119, 8183, 8055, 9609, 10121,
                      12976, 16928, 19089, 23500), # ₹ Crore per month (approximate)
  SIP_Folios_Cr = c(1.00, 1.44, 2.00, 2.50, 3.10, 3.50, 4.20,
                     5.50, 6.80, 8.00, 10.00) # Crore folios
)

p1 <- ggplot(sip_data, aes(x = Year, y = SIP_Monthly_Cr)) +
  geom_col(fill = "#2980b9", alpha = 0.8) +
  geom_line(aes(group = 1), color = "#c0392b", size = 1.2) +
  labs(title = "Monthly SIP Contribution (₹ Crore)", x = "Year", y = "₹ Crore") +
  scale_x_continuous(breaks = 2016:2026) +
  theme_minimal(base_size = 10) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p2 <- ggplot(sip_data, aes(x = Year, y = SIP_Folios_Cr)) +
  geom_col(fill = "#27ae60", alpha = 0.8) +
  geom_line(aes(group = 1), color = "#e67e22", size = 1.2) +
  labs(title = "SIP Folios (Crore)", x = "Year", y = "Crore Folios") +
  scale_x_continuous(breaks = 2016:2026) +
  theme_minimal(base_size = 10) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

gridExtra::grid.arrange(p1, p2, ncol = 2)
```

*Source: Compiled from RBI, SEBI, NSE, BSE, AMFI and Government Reports (Latest Available Data).*

---

## 2.9 Exchange Traded Funds (ETFs) {#etfs}

### 2.9.1 Meaning and Structure

An **Exchange Traded Fund (ETF)** is a mutual fund that is **listed and traded on stock exchanges** like shares, while tracking a specific **index, commodity, or basket of assets**. ETFs combine the diversification benefits of mutual funds with the trading flexibility of stocks.

**Key Features of ETFs:**
- Passively managed (track an index)
- Lower expense ratio than actively managed funds (0.05%–0.50%)
- Real-time pricing (unlike NAV-based mutual funds)
- Can be bought/sold anytime during market hours
- Minimum investment = 1 unit (at market price)
- No exit load in most cases
- Require **demat account** (unlike regular mutual funds)

**ETF vs. Index Fund:**

| Feature | ETF | Index Fund |
|---------|-----|------------|
| Trading | Real-time on exchange | End-of-day NAV |
| Demat | Required | Not required |
| Minimum Amount | 1 unit | ₹100–₹500 SIP |
| Expense Ratio | 0.05%–0.25% | 0.15%–0.50% |
| Tracking Error | Slightly higher | Lower |
| Liquidity | Higher (intraday) | T+3 redemption |

**ETF Growth in India:**

```{r etf-growth, echo=FALSE, fig.cap="Figure 2.7: ETF AUM Growth in India (2016–2026)"}
etf_data <- data.frame(
  Year = 2016:2026,
  ETF_AUM = c(0.41, 0.83, 1.02, 1.49, 2.12, 2.64,
               3.28, 5.16, 6.80, 8.20, 9.50) # ₹ Lakh Crore
)

ggplot(etf_data, aes(x = Year, y = ETF_AUM)) +
  geom_col(fill = "#8e44ad", alpha = 0.8) +
  geom_line(aes(group = 1), color = "#f39c12", size = 1.5) +
  geom_point(color = "#f39c12", size = 3) +
  labs(title = "ETF AUM in India (2016–2026)",
       subtitle = "CAGR ~35% — ETF AUM grew from ₹0.41 to ₹9.50 Lakh Crore",
       x = "Year", y = "AUM (₹ Lakh Crore)") +
  scale_x_continuous(breaks = 2016:2026) +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

*Source: Compiled from RBI, SEBI, NSE, BSE, AMFI and Government Reports (Latest Available Data).*

**Popular ETFs in India (2024):**

| ETF Name | Tracks | AUM (₹ Cr) | Expense Ratio |
|----------|--------|-----------|---------------|
| Nippon India Nifty 50 BeES | Nifty 50 | ~12,000 | 0.04% |
| HDFC Nifty 50 ETF | Nifty 50 | ~10,500 | 0.10% |
| SBI Nifty 50 ETF | Nifty 50 | ~1,10,000+ (EPFO) | 0.07% |
| Nippon India Junior BeES | Nifty Next 50 | ~3,800 | 0.13% |
| ICICI Prudential Midcap ETF | Nifty Midcap 150 | ~800 | 0.18% |
| Nippon India Gold BeES | Gold (MCX) | ~8,500 | 0.82% |
| Nippon India Silver ETF | Silver (MCX) | ~1,200 | 0.65% |
| Bharat Bond ETF April 2030 | PSU Bonds Ladder | ~15,000 | 0.0005% |

*Source: Compiled from RBI, SEBI, NSE, BSE, AMFI and Government Reports (Latest Available Data).*

**Special Feature: EPFO and ETFs**

A landmark development in Indian capital markets was the **EPFO's decision in 2015 to invest 5% (later increased to 15%) of incremental EPF corpus in equity ETFs** tracking Nifty 50 and Sensex. This makes EPFO one of the largest domestic institutional investors through the ETF route, protecting subscribers' retirement savings through passive equity exposure.

---

## 2.10 Real Estate Investment Trusts (REITs) {#reits}

### 2.10.1 Meaning and Structure

A **Real Estate Investment Trust (REIT)** is a listed entity that owns, operates, or finances **income-generating real estate** assets. REITs allow retail investors to invest in large commercial properties (office parks, malls, hospitals) with a minimum of ₹10,000–15,000.

SEBI introduced REIT regulations in **2014**, and the **first Indian REIT — Embassy Office Parks REIT — was listed on NSE in April 2019**.

**Structure of a REIT:**

```{mermaid}
flowchart LR
    A[Retail & Institutional Investors] -->|Units| B[REIT Trust]
    B -->|Manager| C[Investment Manager]
    C -->|Acquires / Manages| D[Special Purpose Vehicles - SPVs]
    D -->|Holds| E[Commercial Properties]
    E --> E1[Office Parks]
    E --> E2[Retail Malls]
    E --> E3[Warehouses]
    E --> E4[Hospitals]
    D -->|Income| B
    B -->|90% Distribution| A
    F[SEBI] -->|Regulates| B
    style B fill:#2c3e50,color:#fff
    style F fill:#c0392b,color:#fff
    style A fill:#2980b9,color:#fff
```

*Figure 2.8: Structure of Indian REITs*

**Mandatory Requirements under SEBI REIT Regulations:**
- **Minimum 80% assets** in completed, rent-generating properties
- **Minimum 90% distributable cash** must be distributed to unit holders every 6 months
- Must be listed on NSE or BSE
- Minimum public float: 25% of total units

**Indian REITs (2024):**

| REIT | Listed | Portfolio | AUM (₹ Cr) | Distribution Yield |
|------|--------|-----------|-----------|-------------------|
| Embassy Office Parks REIT | April 2019 | 45 Mn sqft (Bengaluru, Mumbai, NCR) | ~42,000 | ~6.5% |
| Mindspace Business Parks REIT | August 2020 | 32 Mn sqft (Mumbai, Hyderabad, Pune) | ~25,000 | ~6.2% |
| Brookfield India REIT | February 2021 | 27 Mn sqft (NCR, Mumbai, Kolkata) | ~18,000 | ~7.5% |
| Nexus Select Trust (Retail REIT) | May 2023 | 17 retail malls across India | ~15,000 | ~6.0% |

*Source: Compiled from RBI, SEBI, NSE, BSE, AMFI and Government Reports (Latest Available Data).*

**REITs as an Investment:**
- **Rental income** from commercial properties distributed quarterly
- **Capital appreciation** as property values rise
- Lower entry barrier vs. direct real estate
- Portfolio diversification (real estate exposure without physical ownership)
- **Tax treatment**: REIT distributions partly taxable as business income, partly as dividend

---

## 2.11 Infrastructure Investment Trusts (InvITs) {#invits}

### 2.11.1 Meaning and Structure

An **Infrastructure Investment Trust (InvIT)** is a collective investment vehicle — similar in structure to REITs — that pools investor money to invest in **operational infrastructure projects** such as roads, power transmission lines, gas pipelines, and telecom towers.

SEBI introduced InvIT regulations in **2014**. India's first InvIT — **IRB InvIT Fund** — listed in May 2017.

**REIT vs. InvIT:**

| Feature | REIT | InvIT |
|---------|------|-------|
| Assets | Commercial real estate | Infrastructure projects |
| Revenue Source | Rental income | Toll, transmission charges, gas fees |
| Underlying Asset | Buildings / properties | Roads, pipelines, towers |
| Risk | Lower (stable rents) | Medium (traffic/usage-dependent) |
| First India Listing | Embassy REIT (2019) | IRB InvIT (2017) |

*Source: Compiled from RBI, SEBI, NSE, BSE, AMFI and Government Reports (Latest Available Data).*

**Major Listed InvITs in India (2024):**

| InvIT | Assets | Listed | Yield |
|-------|--------|--------|-------|
| IRB InvIT Fund | National highways (BOT) | May 2017 | ~10–12% |
| IndInfravit Trust | Highway NHAI projects | March 2018 | ~11% |
| Powergrid InvIT | Power transmission assets | May 2021 | ~10% |
| India Grid Trust (IndiGrid) | Power transmission (Sterlite) | June 2017 | ~11–12% |

*Source: Compiled from RBI, SEBI, NSE, BSE, AMFI and Government Reports (Latest Available Data).*

**Growth of REITs and InvITs:**

```{r reit-invit-growth, echo=FALSE, fig.cap="Figure 2.9: REIT and InvIT AUM Growth in India (2017–2026)"}
ri_data <- data.frame(
  Year = 2017:2026,
  REIT_AUM = c(0, 0, 0.10, 0.42, 0.65, 0.90, 1.20, 1.60, 2.10, 2.80),
  InvIT_AUM = c(0.08, 0.20, 0.45, 0.80, 1.10, 1.40, 1.80, 2.20, 2.70, 3.20)
)

ri_long <- tidyr::pivot_longer(ri_data, -Year, names_to = "Type", values_to = "AUM")

ggplot(ri_long, aes(x = Year, y = AUM, fill = Type)) +
  geom_col(position = "dodge", alpha = 0.85) +
  scale_fill_manual(values = c("REIT_AUM" = "#2980b9", "InvIT_AUM" = "#27ae60"),
                    labels = c("REIT AUM", "InvIT AUM")) +
  labs(title = "REIT and InvIT AUM Growth (₹ Lakh Crore)",
       subtitle = "REITs launched 2019; InvITs launched 2017",
       x = "Year", y = "AUM (₹ Lakh Crore)", fill = "") +
  scale_x_continuous(breaks = 2017:2026) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom")
```

*Source: Compiled from RBI, SEBI, NSE, BSE, AMFI and Government Reports (Latest Available Data).*

---

::: {.research-corner}
**Research Corner 2.1: The Future of Passive Investing in India**

The global shift from active to passive investing is accelerating in India. Key research observations:

1. **SPIVA India Scorecard (S&P, 2024)** shows that over 10 years, **over 80% of large-cap active funds underperformed** the Nifty 50 TRI (Total Return Index).

2. **EPFO's ETF mandate** has democratised passive investing, with ₹2.5+ lakh crore invested in equity ETFs by EPFO as of 2024.

3. **Zero-commission direct plans** and growing financial literacy are driving retail investors toward direct index ETFs.

**Research Gap:** The tracking error of Indian ETFs — especially for mid-cap and small-cap indices — remains significantly higher than developed market ETFs. Research into optimal ETF creation/redemption mechanisms for Indian markets is needed.
:::

---

## 2.12 Summary {#unit2-summary}

1. **Shares** are units of company ownership. Equity shareholders are residual claimants with voting rights; preference shareholders have priority in dividend and capital repayment.

2. **Debentures** are debt instruments — NCDs, FCDs, and PCDs represent the primary classifications used in Indian markets. CRISIL, ICRA, and CARE ratings guide investor decisions.

3. **Mutual funds** are the most accessible financial investment vehicle for retail investors — professionally managed, diversified, and SEBI-regulated. India's MF industry crossed ₹67 lakh crore AUM in 2026.

4. **NAV** is the per-unit value of a mutual fund, calculated daily. Total returns depend on NAV appreciation plus dividend distributions.

5. **Expense Ratio** directly erodes investor returns. Direct plans save 0.5%–1.2% annually, compounding to significant wealth over 15–20 years.

6. **SEBI's 2017 categorisation** brought standardisation — one scheme per category per AMC — enabling investors to make meaningful comparisons.

7. **ETFs** track indices passively at very low cost and can be traded intraday. EPFO's ETF mandate is a landmark development for Indian capital markets.

8. **REITs** enable retail investors to invest in commercial real estate with ₹10,000–15,000, earning distribution yields of 6–8% p.a. Four REITs are listed in India (2024).

9. **InvITs** invest in operating infrastructure projects (roads, power lines, gas pipelines), offering yields of 10–12% p.a. with government-backed revenue streams.

---

## Key Terms {#unit2-keyterms}

1. **Share** — Unit of ownership in a company
2. **Equity Share** — Ordinary share with full ownership rights and residual claims
3. **Preference Share** — Share with priority dividend rights over equity
4. **Cumulative Preference Share** — Unpaid dividends accumulate for future payment
5. **Convertible Preference Share** — Convertible into equity shares at specified date
6. **Debenture** — Long-term debt instrument issued by companies
7. **NCD (Non-Convertible Debenture)** — Debenture that cannot be converted to equity
8. **ESOP** — Employee Stock Option Plan; equity compensation for employees
9. **Bonus Share** — Free share issued to existing shareholders from reserves
10. **Rights Issue** — Offer to existing shareholders to buy new shares at discount
11. **Mutual Fund** — Pooled investment vehicle managed by professional AMC
12. **AMC** — Asset Management Company; manages mutual fund investments
13. **NAV** — Net Asset Value; per-unit market value of mutual fund
14. **Expense Ratio / TER** — Annual cost charged by fund as % of AUM
15. **Direct Plan** — Mutual fund plan without distributor commission
16. **Regular Plan** — Mutual fund plan through distributors with higher TER
17. **SIP** — Systematic Investment Plan; regular fixed-amount investment
18. **ELSS** — Equity Linked Savings Scheme; 3-year lock-in, Section 80C benefit
19. **Large Cap Fund** — Fund investing 80%+ in top 100 companies by market cap
20. **Mid Cap Fund** — Fund investing 65%+ in 101st–250th companies by market cap
21. **Gilt Fund** — Fund investing 80%+ in government securities
22. **ETF** — Exchange Traded Fund; passively managed, exchange-listed index fund
23. **Tracking Error** — Deviation of ETF/Index fund returns from benchmark index
24. **REIT** — Real Estate Investment Trust; listed vehicle owning commercial properties
25. **InvIT** — Infrastructure Investment Trust; listed vehicle owning infrastructure assets
26. **Distribution Yield** — Annual distributions as % of unit price (REIT/InvIT)
27. **AMFI** — Association of Mutual Funds in India; industry body
28. **DVR Shares** — Differential Voting Rights shares; promoted by SEBI in 2019
29. **Sweat Equity** — Shares issued to employees/directors for non-cash contributions
30. **EPFO** — Employees' Provident Fund Organisation; invests in equity ETFs

---

## Review Questions — Unit II {#unit2-review}

### Very Short Answer Questions

1. What is the difference between equity share and preference share in one sentence?
2. Name four types of preference shares.
3. What is the full form of ELSS?
4. State the SEBI-mandated distribution percentage for REITs.
5. Who introduced ETFs in India and when was the first equity ETF listed?
6. What is the minimum lock-in period for ELSS?
7. Define Expense Ratio.
8. State the SEBI categorisation rule for large cap funds.
9. Name the four listed REITs in India as of 2024.
10. What is the difference between an NCD and an FCD?

### Essay Questions

1. Explain the SEBI categorisation of mutual fund schemes (2017). Why was this circular significant for the Indian mutual fund industry? Discuss with examples.
2. "Direct plans in mutual funds are always superior to regular plans for long-term investors." Do you agree? Justify with numerical illustrations showing the impact of expense ratio over 20 years.
3. Critically evaluate REITs as an investment avenue for retail investors in India. Discuss the regulatory framework, available REITs, and their historical performance.
4. Explain the structure and growth of InvITs in India. How do they differ from REITs? Which sectors are currently represented in Indian InvITs?
5. Discuss the role of ETFs in democratising equity investment in India. Evaluate the EPFO's ETF mandate and its impact on domestic capital markets.

---

## MCQs — Unit II (Selected)

**1.** The SEBI (Mutual Fund) Regulations were first issued in:
- (A) 1992  (B) 1996 ← **Correct**  (C) 2000  (D) 2003

*Answer: B. SEBI (Mutual Fund) Regulations were enacted in 1996.*

**2.** NAV of a mutual fund is calculated:
- (A) Weekly  (B) Monthly  (C) Daily ← **Correct**  (D) Quarterly

*Answer: C. NAV is disclosed daily on business days by all open-end mutual funds.*

**3.** The first REIT in India was:
- (A) Mindspace REIT  (B) Brookfield REIT  (C) Embassy Office Parks REIT ← **Correct**  (D) Nexus Select Trust

*Answer: C. Embassy Office Parks REIT was listed on NSE in April 2019, the first REIT in India.*

**4.** ELSS funds have a mandatory lock-in of:
- (A) 1 year  (B) 2 years  (C) 3 years ← **Correct**  (D) 5 years

*Answer: C. ELSS has a 3-year lock-in — the shortest among Section 80C instruments.*

**5.** Which of the following is a passively managed fund?
- (A) Flexi Cap Fund  (B) Mid Cap Fund  (C) ELSS Fund  (D) Index Fund ← **Correct**

*Answer: D. Index funds (and ETFs) are passively managed, tracking a benchmark index.*

---

## Numerical Problems — Unit II

**Problem 2.1:** A mutual fund has the following: Equity portfolio ₹5,200 Cr, Bonds ₹1,800 Cr, Cash ₹150 Cr, Receivables ₹50 Cr, Total liabilities ₹80 Cr. Number of units: 350 Cr. Calculate NAV.

```{r nav-problem, echo=TRUE}
equity <- 5200; bonds <- 1800; cash <- 150; recv <- 50
liabilities <- 80; units <- 350

total_assets <- equity + bonds + cash + recv
nav <- (total_assets - liabilities) / units
cat("NAV =", round(nav, 4), "per unit\n")
```

**Problem 2.2:** You invested ₹50,000 in an equity fund at NAV ₹25. After 5 years NAV is ₹56.20. Calculate CAGR.

```{r cagr-problem, echo=TRUE}
nav_buy <- 25; nav_sell <- 56.20; years <- 5
cagr <- (nav_sell/nav_buy)^(1/years) - 1
cat("CAGR:", round(cagr*100, 2), "% per annum\n")
units_held <- 50000/nav_buy
gain <- (nav_sell - nav_buy) * units_held
cat("Investment value:", format(nav_sell * units_held, big.mark = ","), "\n")
cat("Capital Gain: ₹", format(gain, big.mark = ","), "\n")
```

*End of Unit II*

